Bandit Multiclass Linear Classification: \\ Efficient Algorithms for the Separable Case

### 2019-02-06 arxiv/mistake-lower-bound-for-ignorant-algorithms.tex
.

In rounds $1,2,\dots,\tau-1$, the algorithm predicts the incorrect class $2$
with probability at least $q_0$. Thus,
$$
\Exp\left[\sum_{t=1}^{\tau - 1} \indicator{\widehat y_t\neq y_t}\right] = q_0 \Exp[(\tau - 1)] \; .
$$
In rounds $\tau, \tau+1, \dots, T$, all the examples are the same and are equal
to $(v_\tau, 2)$. Let $s$ be the first time step $t$ such that $t \ge \tau$
and the algorithm makes a correct prediction. If the algorithm makes mistakes
in all rounds $\tau, \tau+1, \dots, T$, we define $s = T+1$.
By definition the algorithm makes mistakes in rounds $\tau, \tau+1, \dots, s-1$.
Therefore,
$$
\Exp\left[\sum_{t=\tau}^T \indicator{\widehat y_t\neq y_t}\right] \ge \Exp[s-\tau].
$$
Since the algorithm is ignorant, conditioned on $\tau$, the random variable $s-\tau$
has a truncated geometric distribution with parameter $p_\tau(2|v_\tau) \le q_0$.
Therefore,
$$
\Exp[s-\tau ~|~ \tau]
\ge
\Exp \left[\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right) ~\middle|~ \tau \right] \; .
$$

@@

>>>>

In rounds $1,2,\dots,\tau-1$, the algorithm predicts the incorrect class $2$
with probability at least $q_0$. Thus,
$$
\Exp\left[\sum_{t=1}^{\tau - 1} \indicator{\widehat y_t\neq y_t}\right] = q_0 \Exp[(\tau - 1)] \; .
$$
In rounds $\tau, \tau+1, \dots, T$, all the examples are the same and are equal
to $(v_\tau, 2)$. Let $s$ be the first time step $t$ such that $t \ge \tau$
and the algorithm makes a correct prediction. If the algorithm makes mistakes
in all rounds $\tau, \tau+1, \dots, T$, we define $s = T+1$.
By definition the algorithm makes mistakes in rounds $\tau, \tau+1, \dots, s-1$.
Therefore,
$$
\Exp\left[\sum_{t=\tau}^T \indicator{\widehat y_t\neq y_t}\right] \ge \Exp[s-\tau].
$$
%Since the algorithm is ignorant, conditioned on $\tau$, $s-\tau+1$ has the same distribution as
%$\min(X, T-\tau+2)$, where $X$ is drawn from a geometric distribution with parameter
%$p_\tau(2|v_\tau)$.
Since the algorithm is ignorant, conditiond on $\tau$, $s-\tau$ follows a truncated 
geometric distribution with parameter $q\triangleq p_\tau(2|v_\tau)$
(i.e., $s-\tau$ is $0$ with probability $q$, $1$ with probability $(1-q)q$, $2$ with probability $(1-q)^2q, \ldots$). Its expectation can be calculated as follows: 
\begin{align*}
\Exp[s-\tau~|~\tau] 
&= \sum_{i=1}^{T+1-\tau} i\times  \Pr[s-\tau=i]= \sum_{i=1}^{T+1-\tau}\sum_{j=1}^i \Pr[s-\tau=i] = \sum_{j=1}^{T+1-\tau}\sum_{i=j}^{T+1-\tau} \Pr[s-\tau=i] \\
&= \sum_{j=1}^{T+1-\tau}\Pr[s-\tau\geq j] = \sum_{j=1}^{T+1-\tau}(1-q)^j \geq  \sum_{j=1}^{T+1-\tau}(1-q_0)^j = \frac{1-q_0}{q_0}\left( 1-(1-q_0)^{T-\tau+1} \right). 
\end{align*}
%By Lemma~\ref{lem:truncated-geom} below,
%\[
%  \Exp\left[s-\tau+1 \mid \tau, p_\tau(2|v_\tau) \right] = F(p_\tau(2|v_\tau), T-\tau+2) \geq F(q_0, T-\tau+2).
%\]
%Therefore,
%$$
%\Exp[s-\tau ~|~ \tau]
%\ge
%F(q_0, T-\tau+2) - 1
%=
%\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right) \; .
%$$

%the random variable $s-\tau+1$
%has a truncated geometric distribution with parameter $p_\tau(2|v_\tau) \le q_0$.
%Specifically, conditioned on $\tau$,

@@

%\begin{lemma}
%For $p \in (0,1)$ and $N \geq 1$, denote by $F(p, N) = \frac{1}{p}(1 - (1-p)^N)$.
%\begin{enumerate}
%  \item $F(p, N)$ is monotonically decreasing in $p$.
%  \item If $X \sim \Geom(p)$, then $\Exp[\min(X, N)] = F(p, N)$.
%\end{enumerate}
%\label{lem:truncated-geom}
%\end{lemma}
%\begin{proof}
%We show the two items respectively.
%\begin{enumerate}
%\item By calculus,
%\[ \frac{\partial F(p,n)}{\partial p} = \frac{(1-p)^{N-1} (1 + (N-1)p) - 1}{p^2}. \]
%Now, note that for any $p \in (0,1)$,
%\[ (1 + (N-1)p) \leq \frac{1}{(1 - p)^{N-1}} \]
%We therefore get $\frac{\partial F(p,n)}{\partial p} < 0$ for all $p$, showing the first item.

%\item We have the following decomposition on $\Exp[X]$:
%\[ \Exp[X] = \Exp[X \one[X \leq N]] + \Prob(X \geq N+1) \Exp[X|X \geq N+1] \]
%By the memoryless property of geometric distribution, $\Exp[X|X \geq N+1] = N + \Exp[X] = N + \frac 1 p$.
%Therefore,
%\[ \Exp[X] = \Exp[X \one[X \leq N]] + \Prob(X \geq N+1) \cdot N + \Prob(X \geq N+1) \cdot \frac 1 p. \]
%At the same time, observe that $\Exp[X \one[X \leq N]] + \Prob(X \geq N+1) \cdot N = \Exp[\min(X, N)]$.
%Therefore,
%\[ \Exp[\min(X, N)] = \frac 1 p \cdot (1 - \Prob(X \geq N+1)) = \frac{1}{p}(1 - (1-p)^N) = F(p, N). \qedhere \]
%\end{enumerate}
%\end{proof}

### 2019-02-06 arxiv/np-hardness-of-the-weak-labeling-problem.tex
widened the eqns
.

We use symbols $1,2,\dots,K$ for strong labels and symbols $\overline{1},
\overline{2}, \dots, \overline{K}$ for weak labels. We adopt the convention that
$\overline{\overline{i}} = i$ for any symbol $i$. Formally, the weak
labeling problem can be described as below:
\begin{figure}[H]
\begin{framed}
\begin{center}
    \textbf{Weak Labeling}
\end{center}
\textbf{Given:} Feature-label pairs $(x_1, y_1)$, $(x_2, y_2)$, \dots, $(x_T, y_T)$ in $\{0,1\}^d \times \{1,2,\dots, K, \overline{1}, \overline{2}, \dots, \overline{K}\}$. \\
\textbf{Question:} Do there exist $ w_1, w_2, \dots, w_K \in \R^d$ such that for all $t=1,2,\dots,T$,
\begin{align*}
& y_t \in \{1,2,\dots,K\} \Longrightarrow  \\
& \quad \forall i \in \{1,2,\dots,K\} \setminus \{y_t\} \quad \ip{w_{y_t}}{x_t}  > \ip{w_i}{x_t} \; , \\
& \text{and} \\
& y_t \in \{\overline{1}, \overline{2},\dots, \overline{K}\} \Longrightarrow \\
& \quad \exists i \in \{1,2,\dots,K\} \quad \ip{w_i}{x_t} > \ip{w_{\overline{y_t}}}{x_t} \; ?
\end{align*}
\end{framed}
\end{figure}

@@

Below we show the reduction. Suppose we are given an instance of the set
splitting problem
\begin{align*}
S & = \{1, 2, \dots, N\} \; , \\
C & = \{c_1, c_2, \dots, c_M\} \; .
\end{align*}

@@

For example, if we have $S=\{1,2,3\}$, $C=\{c_1, c_2\}$, $c_1 = \{1,2\}$,
$c_2=\{2,3\}$, then we create the weak labeling sample set as:
\begin{multline*}
\{
((0,0,0,1),3), ((1,0,0,1),\overline{3}), ((0,1,0,1),\overline{3}), \\
((0,0,1,1),\overline{3}), ((1,1,0,1),3), ((0,1,1,1),3)
\} \; .
\end{multline*}
The following lemma shows that answering this weak labeling problem is
equivalent to answering the original set splitting problem.

@@

\begin{proof}
$(\Longrightarrow)$ Let $S_1, S_2$ be the solution of the set splitting problem. Define
$$
w_1 = \left(a_1, a_2, \cdots, a_N, -\frac{1}{2}\right),
$$
where for all $i \in \{1,2,\dots,N\}$, $a_i=1$ if $i\in S_1$ and $a_i=-N$ if
$i\notin S_1$. Similarly, define
$$
w_2 = \left(b_1, b_2, \cdots, b_N, -\frac{1}{2}\right),
$$
where for all $i \in \{1,2,\dots,N\}$, $b_i=1$ if $i \in S_2$ and $b_i=-N$ if
$i\notin S_2$. Finally, define
$$
w_3 = (0,0,\cdots, 0),
$$
the zero vector. To see this is a solution for the weak labeling problem, we
verify separately for Type 1-3 samples defined above. For Type 1 sample, we have
$$
\ip{w_3}{x} = 0 > -\frac{1}{2} = \ip{w_1}{x}=\ip{w_2}{x}.
$$
For a Type 2 sample that corresponds to index $i$, we have either $i\in S_1$ or
$i\in S_2$ because $S_1\cup S_2 = \{1,2,\dots,N\}$ is guaranteed. Thus, either
$a_i=1$ or $b_i=1$. If $a_i=1$ is the case, then
$$
\ip{w_1}{x} = a_i - \frac{1}{2} = \frac{1}{2} > 0 = \ip{w_3}{x};
$$
similarly if $b_i=1$, we have $\ip{w_2}{x}>\ip{w_3}{x}$. \\ For a Type 3 sample
that corresponds to index $j$, Since $c_j \not\subset S_1$, there exists some
$i'\in c_j$ and $i'\notin S_1$. Thus we have $x_{i'}=1$, $a_{i'}=-N$, and
therefore
\begin{align*}
\ip{w_1}{x}
& = a_{i'}x_{i'} + \sum_{i\in \{1,2,\dots,N\} \setminus \{i'\}} a_ix_i - \frac{1}{2} \\
& \le -N + (N-1)-\frac{1}{2} \\
& < 0  \\
& = \ip{w_3}{x} \; .
\end{align*}
Because $c_j \not\subset S_2$ also holds, we also have
$\ip{w_2}{x}<\ip{w_3}{x}$. This direction is therefore proved. \\
\ \\
$(\Longleftarrow)$ Given the solution $w_1, w_2, w_3$ of the weak labeling problem, we define
\begin{align*}
S_1 &= \left\{i \in \{1,2,\dots,n\} ~:~ \ip{w_1-w_3}{(\e_i, 1)} > 0 \right\}, \\
S_2 &= \left\{i \in \{1,2,\dots,n\} ~:~ \ip{w_2-w_3}{(\e_i, 1)} > 0 \right. \\
& \qquad \qquad \qquad \left. \text{\ and\ } i \notin S_1 \right\}.
\end{align*}
It is not hard to see $S_1 \cap S_2 = \emptyset$ and $S_1\cup S_2 =
\{1,2,\dots,N\}$. The former is because $S_2$ only includes elements that are
not in $S_1$. For the latter, note that $(\e_i, 1)$ is the feature vector for
Type 2 samples. Because Type 2 samples all have label $\overline{3}$, for any $i
\in \{1,2,\dots,N\}$, one of the following must hold: $\ip{w_1-w_3}{(\e_i,
1)}>0$ or $\ip{w_2-w_3}{(\e_i, 1)}>0$. This implies $i\in S_1$ or $i\in S_2$.

Now we show $\forall j$, $c_j \not\subset S_1$ and $c_j \not\subset S_2$ by
contradiction. Assume there exists some $j$ such that $c_j \subset S_1$. By our
definition of $S_1$, we have $\ip{w_1-w_3}{(\e_i, 1)} > 0$ for all $i\in c_j$.
Therefore,
\begin{align*}
\sum_{i\in c_j} \ip{w_1-w_3}{\left(\e_i, 1\right)}
& = \ip{w_1-w_3}{\left(\sum_{i\in c_j} \e_i, |c_j|\right)}  \\
& > 0.
\end{align*}
Because Type 1 sample has label $3$, we also have
$$
\ip{w_1-w_3}{\left(\zero, 1\right)} < 0.
$$
Combining the above two inequalities, we get
\begin{align*}
& \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, 1\right)} \\
& = \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, |c_j|\right)}  \\
& \qquad - (|c_j|-1)\ip{w_1-w_3}{\left(\zero, 1\right)} \\
& > 0 \; .
\end{align*}
Note that $\left(\sum_{i\in c_j}\e_i, 1\right)$ is a feature vector for Type 3
samples. Thus the above inequality contradicts that Type 3 samples have label 3.
Therefore, $c_j \not\subset S_1$. If we assume there exists some $c_j \subset
S_2$, same arguments apply and also lead to contradiction.
\end{proof}

>>>>

We use symbols $1,2,\dots,K$ for strong labels and symbols $\overline{1},
\overline{2}, \dots, \overline{K}$ for weak labels. Formally, the weak
labeling problem can be described as below:
\begin{figure}[H]
\begin{framed}
\begin{center}
    \textbf{Weak Labeling}
\end{center}
\textbf{Given:} Feature-label pairs $(x_1, y_1)$, $(x_2, y_2)$, \dots, $(x_T, y_T)$ in $\{0,1\}^d \times \{1,2,\dots, K, \overline{1}, \overline{2}, \dots, \overline{K}\}$. \\
\textbf{Question:} Do there exist $ w_1, w_2, \dots, w_K \in \R^d$ such that for all $t=1,2,\dots,T$,
\begin{align*}
& y_t \in \{1,2,\dots,K\} \Longrightarrow \forall i \in \{1,2,\dots,K\} \setminus \{y_t\} \quad \ip{w_{y_t}}{x_t}  > \ip{w_i}{x_t} \; , \\
& \text{and} \\
& y_t \in \{\overline{1}, \overline{2},\dots, \overline{K}\} \Longrightarrow \exists i \in \{1,2,\dots,K\} \quad \ip{w_i}{x_t} > \ip{w_{\overline{y_t}}}{x_t} \; ?
\end{align*}
\end{framed}
\end{figure}

@@

Below we show the reduction. Suppose we are given an instance of the set
splitting problem
\begin{align*}
S = \{1, 2, \dots, N\} \; ,
C = \{c_1, c_2, \dots, c_M\} \; .
\end{align*}

@@

For example, if we have $S=\{1,2,3\}$, $C=\{c_1, c_2\}$, $c_1 = \{1,2\}$,
$c_2=\{2,3\}$, then we create the weak labeling sample set as:
\[
\{
((0,0,0,1),3), ((1,0,0,1),\overline{3}), ((0,1,0,1),\overline{3}),
((0,0,1,1),\overline{3}), ((1,1,0,1),3), ((0,1,1,1),3)
\} \; .
\]
The following lemma shows that answering this weak labeling problem is
equivalent to answering the original set splitting problem.

@@

\begin{proof}
$(\Longrightarrow)$ Let $S_1, S_2$ be the solution of the set splitting problem. Define
$$
w_1 = \left(a_1, a_2, \cdots, a_N, -\frac{1}{2}\right),
$$
where for all $i \in \{1,2,\dots,N\}$, $a_i=1$ if $i\in S_1$ and $a_i=-N$ if
$i\notin S_1$. Similarly, define
$$
w_2 = \left(b_1, b_2, \cdots, b_N, -\frac{1}{2}\right),
$$
where for all $i \in \{1,2,\dots,N\}$, $b_i=1$ if $i \in S_2$ and $b_i=-N$ if
$i\notin S_2$. Finally, define
$$
w_3 = (0,0,\cdots, 0),
$$
the zero vector. To see this is a solution for the weak labeling problem, we
verify separately for Type 1-3 samples defined above. For Type 1 sample, we have
$$
\ip{w_3}{x} = 0 > -\frac{1}{2} = \ip{w_1}{x}=\ip{w_2}{x}.
$$
For a Type 2 sample that corresponds to index $i$, we have either $i\in S_1$ or
$i\in S_2$ because $S_1\cup S_2 = \{1,2,\dots,N\}$ is guaranteed. Thus, either
$a_i=1$ or $b_i=1$. If $a_i=1$ is the case, then
$$
\ip{w_1}{x} = a_i - \frac{1}{2} = \frac{1}{2} > 0 = \ip{w_3}{x};
$$
similarly if $b_i=1$, we have $\ip{w_2}{x}>\ip{w_3}{x}$. \\ For a Type 3 sample
that corresponds to index $j$, Since $c_j \not\subset S_1$, there exists some
$i'\in c_j$ and $i'\notin S_1$. Thus we have $x_{i'}=1$, $a_{i'}=-N$, and
therefore
\begin{align*}
\ip{w_1}{x}
& = a_{i'}x_{i'} + \sum_{i\in \{1,2,\dots,N\} \setminus \{i'\}} a_ix_i - \frac{1}{2} \\
& \le -N + (N-1)-\frac{1}{2} < 0 = \ip{w_3}{x} \; .
\end{align*}
Because $c_j \not\subset S_2$ also holds, we also have
$\ip{w_2}{x}<\ip{w_3}{x}$. This direction is therefore proved. \\
\ \\
$(\Longleftarrow)$ Given the solution $w_1, w_2, w_3$ of the weak labeling problem, we define
\begin{align*}
S_1 &= \left\{i \in \{1,2,\dots,n\} ~:~ \ip{w_1-w_3}{(\e_i, 1)} > 0 \right\}, \\
S_2 &= \left\{i \in \{1,2,\dots,n\} ~:~ \ip{w_2-w_3}{(\e_i, 1)} > 0 \text{\ and\ } i \notin S_1 \right\}.
\end{align*}
It is not hard to see $S_1 \cap S_2 = \emptyset$ and $S_1\cup S_2 =
\{1,2,\dots,N\}$. The former is because $S_2$ only includes elements that are
not in $S_1$. For the latter, note that $(\e_i, 1)$ is the feature vector for
Type 2 samples. Because Type 2 samples all have label $\overline{3}$, for any $i
\in \{1,2,\dots,N\}$, one of the following must hold: $\ip{w_1-w_3}{(\e_i,
1)}>0$ or $\ip{w_2-w_3}{(\e_i, 1)}>0$. This implies $i\in S_1$ or $i\in S_2$.

Now we show $\forall j$, $c_j \not\subset S_1$ and $c_j \not\subset S_2$ by
contradiction. Assume there exists some $j$ such that $c_j \subset S_1$. By our
definition of $S_1$, we have $\ip{w_1-w_3}{(\e_i, 1)} > 0$ for all $i\in c_j$.
Therefore,
\begin{align*}
\sum_{i\in c_j} \ip{w_1-w_3}{\left(\e_i, 1\right)}
 = \ip{w_1-w_3}{\left(\sum_{i\in c_j} \e_i, |c_j|\right)} > 0.
\end{align*}
Because Type 1 sample has label $3$, we also have
$$
\ip{w_1-w_3}{\left(\zero, 1\right)} < 0.
$$
Combining the above two inequalities, we get
\begin{align*}
& \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, 1\right)}
 = \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, |c_j|\right)} - (|c_j|-1)\ip{w_1-w_3}{\left(\zero, 1\right)}  > 0 \; .
\end{align*}
Note that $\left(\sum_{i\in c_j}\e_i, 1\right)$ is a feature vector for Type 3
samples. Thus the above inequality contradicts that Type 3 samples have label 3.
Therefore, $c_j \not\subset S_1$. If we assume there exists some $c_j \subset
S_2$, same arguments apply and also lead to contradiction.
\end{proof}
